Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.

### 2019-03-26 _book/electoral-modelling-draft.tex
update cache

Joining these two data sources is problematic as there in an inherent asynchronicity in the two events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of obtaining appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see \citet{DavisStimson98}, \citet{Stimson06}, \citet{Liao09} and \citet{Stimson09}). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are imputed using a spatio-temporal imputation that combines areal interpolation \citep{Goodchild1993} and linear time-interpolation. The process of collecting and wrangling the raw data is outlined in section \ref{data}, and the imputation process is detailed in section \ref{imputation}. All data and associated documentation relating to this procedure are available in the \texttt{eechidna} \texttt{R} package \citep{eechidna}, providing a resource for future analysis.

@@

\hypertarget{data}{%
\chapter{Data collection}\label{data}}

\hypertarget{voting-data}{%
\section{Voting data}\label{voting-data}}

@@

\hypertarget{census-data}{%
\section{Census data}\label{census-data}}

@@

\hypertarget{downloading-and-wrangling}{%
\section{Downloading and wrangling}\label{downloading-and-wrangling}}

@@

\hypertarget{imputation}{%
\chapter{Joining Census and election data}\label{imputation}}

\hypertarget{differences-between-census-and-election-data}{%
\section{Differences between Census and election data}\label{differences-between-census-and-election-data}}

@@

\hypertarget{spatio-temporal-imputation}{%
\section{Spatio-temporal imputation}\label{spatio-temporal-imputation}}

@@

\hypertarget{applied}{%
\section{Applied}\label{applied}}

@@

Table 4.1 details the resultant estimated model coefficients and their estimated standard errors for each of the six elections. These are interpreted in the next section.

@@

\begin{center}\includegraphics{electoral-modelling-draft_files/figure-latex/rhovis-1} \end{center}

The spatial autoregressive coefficient \(\rho\) is positive and significant in only the 2001 and 2016 elections, meaning that in these elections, an electorate's political climate was affected by the attitudes of it's neighbours. Conversely, in the other four elections, it appears that after accounting for electoral socio-demographics, there is no spatial effect, meaning electorates effectively acted independently.

@@

To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used. These show the direction, size and significance of an estimated effect, as the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95\% confidence band. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. Plots for each election are faceted to compare the effects over time.

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Although topical in the lead up to an election, incomes only significantly affected electoral two-party preference in 2016 (see Figure \ref{fig:plotincomes}), with higher income areas associated with support for the Liberal party. In the prior elections, the estimated effect is still positive, albeit insignificant. Unemployment on the other hand, was influential in 2001, 2004 and 2007 elections(Figure \ref{fig:plotunemploy}), with electorates with higher unemployment favouring Labor. This estimated effect has become weaker over time, with unemployment having virtually no effect in 2016.

@@

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotincomes-1}

@@

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as \texttt{Extractive} industries) are consistently linked with higher support for the Liberal party, with the mangitude of this effect slightly increasing over the years (see Figure \ref{fig:plotextractive}). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (\texttt{Transformative}), as well as wholesale trade, retail trade, transport, post or warehousing (\texttt{Distributive}), are also linked with the Liberal party. However, these effects are not as strong, and \texttt{Transformative} is marginally insignificant in the 2010 election, whereas \texttt{Distributive} was not a key socio-demographic in 2001 or 2016. These effects are shown in Figures \ref{fig:plottransformative} and \ref{fig:plotdistributive}.

@@

Electorates with a higher proportion of people that have recently moved house (meaning in last five years) are more likely to support the Liberal party, with this effect becoming stronger since the 2010 election. Having controlled for characteristics of house ownership and rental prices (via the factors \texttt{PropertyOwned} and \texttt{RentLoan} respectively), this is effect is somewhat surprising. \emph{Add more here?}

@@

\emph{Potentially remove plots}

\hypertarget{residuals}{%
\section{Residuals}\label{residuals}}

Largest residuals

Residuals by state

\begin{center}\includegraphics{electoral-modelling-draft_files/figure-latex/unnamed-chunk-2-1} \end{center}

>>>>

Joining these two data sources is problematic as there in an inherent asynchronicity in the two events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of obtaining appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see \citet{DavisStimson98}, \citet{Stimson06}, \citet{Liao09} and \citet{Stimson09}). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are imputed using a spatio-temporal imputation that combines areal interpolation \citep{Goodchild1993} and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in section \ref{data}. All data and associated documentation relating to this procedure are available in the \texttt{eechidna} \texttt{R} package \citep{eechidna}, providing a resource for future analysis .

@@

\hypertarget{data}{%
\chapter{Data collection, wrangling and imputation}\label{data}}

\hypertarget{collecting-the-data}{%
\section{Collecting the data}\label{collecting-the-data}}

@@

@@

@@

\hypertarget{joining-census-and-election-data}{%
\section{Joining Census and election data}\label{joining-census-and-election-data}}

\hypertarget{differences-between-census-and-election-data}{%
\subsection{Differences between Census and election data}\label{differences-between-census-and-election-data}}

@@

\hypertarget{spatio-temporal-imputation}{%
\subsection{Spatio-temporal imputation}\label{spatio-temporal-imputation}}

@@

\hypertarget{applied}{%
\subsection{Applied}\label{applied}}

@@

Table 3.1 details the resultant estimated model coefficients and their estimated standard errors for each of the six elections. These are interpreted in the next section.

@@

\textbackslash begin\{figure\}{[}h{]}

\{\centering \includegraphics[height=0.15\textheight]{electoral-modelling-draft_files/figure-latex/rhovis-1}

\}

\textbackslash caption\{Estimates of the spatial autoregressive parameter for each of the six elections, with a 95\% confidence interval. Only in 2001 and 2016 is there a significant spatial component.\}\label{fig:rhovis}
\textbackslash end\{figure\}

The spatial autoregressive coefficient \(\rho\) is positive and significant in only the 2001 and 2016 elections, meaning that in these elections, an electorate's political climate was affected by the attitudes of it's neighbours. Conversely, in the other four elections, it appears that after accounting for electoral socio-demographics, there is no significant spatial effect, meaning electorates effectively acted independently.

@@

To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used. These show the direction, size and significance of an estimated effect - the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95\% confidence band. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. Plots for each election are faceted to compare the effects over time.

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Despite this, incomes have only significantly affected the electoral two-party preference in 2016 (see Figure \ref{fig:plotincomes}), with higher income areas associated with support for the Liberal party. In the prior elections, the estimated effect is still positive, albeit insignificant. Unemployment on the other hand, was influential in 2001, 2004 and 2007 elections (Figure \ref{fig:plotunemploy}), with electorates with higher unemployment favouring Labor. This estimated effect has become weaker over time, with unemployment having virtually no effect in 2016.

@@

{\centering \includegraphics[width=1\linewidth]{electoral-modelling-draft_files/figure-latex/plotincomes-1}

@@

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as \texttt{Extractive} industries) are consistently linked with higher support for the Liberal party, with the mangitude of this effect slightly increasing over the years (see Figure \ref{fig:plotextractive}). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (\texttt{Transformative}), as well as wholesale trade, retail trade, transport, post or warehousing (\texttt{Distributive}), are also linked with the Liberal party. However, these effects are not as strong, and \texttt{Transformative} was marginally insignificant in the 2010 election, whereas \texttt{Distributive} was not a key socio-demographic in 2001 or 2016. These effects are shown in Figures \ref{fig:plottransformative} and \ref{fig:plotdistributive}.

@@

In each of the six elections, electorates with a higher proportion of people that have recently moved house (meaning in last five years) were more likely to support the Liberal party, with this effect becoming stronger since 2010. Having controlled for characteristics of house ownership and rental prices (via the factors \texttt{PropertyOwned} and \texttt{RentLoan} respectively), this is effect is somewhat surprising. \emph{Add more here?}

@@

\hypertarget{a-note-on-similar-variables}{%
\subsection{A note on similar variables}\label{a-note-on-similar-variables}}

Many of the Census variables represent similar information, which is why factors were created and some variables were removed. However, there still remain some variables which are closely related. For example, \texttt{OtherLanguageHome} is the proportion of households that speak a language other than English at home, which is likely to be related to the proportion of people born overseas (a sum of the variables with a \texttt{Born} prefix). In 2001, the coefficient estiamte for \texttt{OtherLanguageHome} is significant and negative, and all \texttt{Born} variables are insignificant. If \texttt{OtherLanguageHome} is removed from the model, populations born in South Eastern Europe (\texttt{Born\_SE\_Europe}) absorb the negative effect (\(p = 0.0011\)).

\hypertarget{a-closer-look-at-the-residuals}{%
\section{A closer look at the residuals}\label{a-closer-look-at-the-residuals}}

\hypertarget{residuals-by-state}{%
\subsection{Residuals by state}\label{residuals-by-state}}

It is often hypothesized that states have systematic differences that cause their electorates to vote differently. Boxplots of residuals grouped by state reveal that only Tasmania and the Australian Capital Territory appear to have a state-specific effect that is not captured by the models. These states appear to have a bias towards Labor. There are few electorates in these states (five and two, respectively), so this might be due to incumbent effects rather than an actual state-specific bias.

{\centering \includegraphics[width=0.75\linewidth]{electoral-modelling-draft_files/figure-latex/resstate-1}

\caption{Boxplot of residuals by state.}\label{fig:resstate}
\end{figure}

\hypertarget{electorates-with-largest-residuals}{%
\subsection{Electorates with largest residuals}\label{electorates-with-largest-residuals}}

The electorates Capricornia (QLD), Leichhardt (QLD) and Riverina (NSW) have the three largest maximum absolute residuals across the six elections. Capricornia is a swing electorate, which strongly supported Labor between 2001 and 2010, before being narrowly won by Liberal in the last two elections. It has many people working in extractive jobs, an above-average Christian population, and above-average proportion of de facto relationships. The model tends to overpredict its two-party preferred vote (in favour of the Liberal party). Leichhardt is an area encompassing the northernmost tip of Queensland and consists of mainly small indigenous communities and uninhabited rural areas, which was won by the Liberal party in all but the 2007 election. It too has an above-average proportion of de facto relationships, along with an above-average proportion of people moving houses and a high proportion of single parent households. The model consistently underpredicts its two-party vote. Riverina is a rural electorate of NSW, and has been a very safe Liberal (National) seat since 1980. It has a relatively monocultural population consisting of a high proportion of Australian citizens and Christians, an above-average amount of extractive jobs and below average education levels. Its Liberal preference is also underpredicted.

\hypertarget{concluding-comments}{%
\chapter{Concluding comments}\label{concluding-comments}}

Across the past six elections, the spatial error models reveal that the electoral socio-demographics driving the two-party preferred vote have remained relatively steady, with industry and type of work being particularly influential, along with the type of relationships and household mobility, among others. The neighbourhood (spatial) effects were only found to be significant in the 2001 and 2016 elections. Surprisingly, unemployment, incomes and family/household sizes were significant in only some elections, and other characteristics like education were not significant at all.

The cross-sectional data set of Census and election data were obtained by collecting information from the ABS and AEC, and then applying a spatio-temporal imputation process, as detailed in section \ref{data}. This data is available in the \texttt{eechidna} \texttt{R} package, which can be downloaded from CRAN. All associated code to produce the data in the package from raw files can be found in the github repository (\url{https://github.com/ropenscilabs/eechidna}), and vignettes on how to use the data are located on the github page (\url{https://ropenscilabs.github.io/eechidna/}).

Furthermore, this paper is generated using the \texttt{knitr} \citep{knitr} package, so code to reproduce the results are contained in the \texttt{RMarkdown} file.

\hypertarget{software}{%
\chapter{Software}\label{software}}

\texttt{eechidna} (Exploring Election and Census Highly Informative Data Nationally for Australia) is an R package that makes it easy to look at the data from the Australian Federal elections and Censuses that occurred between 2001 and 2016. The data in this package includes voting results for each polling booth and electoral division (electorate), Census information for each electorate and maps of the electorates that were in place at the time of each event. All data in this package is obtained from the Australian Electoral Commission, the Australian Bureau of Statistics and the Australian Government. Census data is imputed for the election years in which a Census does not occur via a spatio-temporal imputation method.

The Census and election datasets in the package can be linked by electorate using the unique identifier variable \texttt{UniqueID}. Vignettes describing how to use the data and how the data was obtained, can also be found on \url{https://ropenscilabs.github.io/eechidna/}.
